---
title: "project 4"
author: "Deli Yang; Tianhong Yin"
date: "2022-12-07"
output: html_document
---

## System I: Recommendation based on genres

### Suppose you know the user's favorite genre. How would you recommend movies to him/her? Propose two recommendation schemes along with all necessary technical details

```{r}
# seed and libraries
set.seed(1360)
library(dplyr)
library(ggplot2)
library(recommenderlab)
library(DT)
library(data.table)
library(reshape2)
```

### Exploratory Data Analysis

```{r}
# download and read data
myurl = "https://liangfgithub.github.io/MovieData/"

# ratings data
# use colClasses = 'NULL' to skip columns
ratings = read.csv(paste0(myurl, 'ratings.dat?raw=true'),
                   sep = ':',
                   colClasses = c('integer', 'NULL'),
                   header = FALSE)
colnames(ratings) = c('UserID', 'MovieID', 'Rating', 'Timestamp')

# movies data
# in movies.dat, some movie names contain single colon (:), so the method above does not work
movies = readLines(paste0(myurl, 'movies.dat?raw=true'))
movies = strsplit(movies, split = "::", fixed = TRUE, useBytes = TRUE)
movies = matrix(unlist(movies), ncol = 3, byrow = TRUE)
movies = data.frame(movies, stringsAsFactors = FALSE)
colnames(movies) = c('MovieID', 'Title', 'Genres')
movies$MovieID = as.integer(movies$MovieID)
# convert accented characters
movies$Title[73]
movies$Title = iconv(movies$Title, "latin1", "UTF-8")
movies$Title[73]
# extract year
movies$Year = as.numeric(unlist(
  lapply(movies$Title, function(x) substr(x, nchar(x)-4, nchar(x)-1))))

# user data
users = read.csv(paste0(myurl, 'users.dat?raw=true'),
                 sep = ':', header = FALSE)
users = users[, -c(2,4,6,8)] # skip columns
colnames(users) = c('UserID', 'Gender', 'Age', 'Occupation', 'Zip-code')

dim(users)
length(unique(ratings$UserID))
```

```{r}
# Distribution of ratings: most of the ratings are in the 3-5 range
tmp = data.frame(Rating = 1:5, 
                 freq = as.vector(table(ratings$Rating)/nrow(ratings)))

ggplot(data = tmp, aes(x = Rating, y = freq)) +
  geom_bar(stat="identity", fill = 'steelblue', width = 0.6) + 
  geom_text(aes(label=round(freq, dig=2)), 
                vjust=1.6, color="white", size=3.5) +
  theme_minimal()
```

```{r}
# Ratings per user: most users have left less than 500 ratings 
tmp = ratings %>% 
  group_by(UserID) %>% 
  summarize(ratings_per_user = n()) 

summary(tmp$ratings_per_user) # one has 2314 ratings
sort(tmp$ratings_per_user[tmp$ratings_per_user>1300]) # about 9 users have had over 1300 ratings

tmp %>%
  ggplot(aes(ratings_per_user)) +
  geom_bar(fill = "steelblue") + coord_cartesian(c(20, 500))
```

```{r}
# Ratings per movie
tmp = ratings %>% 
  group_by(MovieID) %>% 
  summarize(ratings_per_movie = n(), ave_ratings = mean(Rating)) %>%
  inner_join(movies, by = 'MovieID')

summary(tmp$ave_ratings) # the highest average rating is 5, which is the maximum, and the lowest is 1. Mean average rating is 3.239.

summary(tmp$ratings_per_movie)

tmp %>% 
  filter(ratings_per_movie > 2000) %>%
  arrange(desc = ratings_per_movie) %>%
  select(c("Title", "ratings_per_movie")) %>%
  print(n = 31)

# The most popular movie is "American Beauty (1999)", which is the last one in the following result, which is followed by the “Star Wars” series. Throughout, "popular" means receiving many ratings. A popular movie may not be a highly-rated movie.
tmp %>% 
  filter(ratings_per_movie > 2000) %>%
  arrange(desc = ratings_per_movie) %>%
  select(c("Title", "ratings_per_movie")) %>%
  print(n = 31)
```

```{r}
# create a binary indicator for the 18 genres for each movie
genres = as.data.frame(movies$Genres, stringsAsFactors=FALSE)
tmp = as.data.frame(tstrsplit(genres[,1], '[|]',
                              type.convert=TRUE),
                    stringsAsFactors=FALSE)
genre_list = c("Action", "Adventure", "Animation",
               "Children's", "Comedy", "Crime",
               "Documentary", "Drama", "Fantasy",
               "Film-Noir", "Horror", "Musical",
               "Mystery", "Romance", "Sci-Fi",
               "Thriller", "War", "Western")
m = length(genre_list)

############## produce the the genre_matrix loaded at the beginning of server.R file
genre_matrix = matrix(0, nrow(movies), length(genre_list))
for(i in 1:nrow(tmp)){
  genre_matrix[i,genre_list %in% tmp[i,]]=1
} 

colnames(genre_matrix) = genre_list
save(genre_matrix, file = "genre_matrix.rda") ## the genre_matrix.rda file
####################################################################################

remove("tmp", "genres")

# combine genre_matrix and movie id
genre_com = as.data.frame(genre_matrix)
genre_com = cbind(movies$MovieID, genre_com)
colnames(genre_com) = c("MovieID", genre_list)
genre_com = as_tibble(genre_com)
```

```{r}
# histograms of the 18 genres based on movies: about half of the movies belong to only the genre drama 
data.frame(Genres = genre_list, 
                 Freq = as.vector(colMeans(genre_matrix))) %>% 
  ggplot(aes(reorder(Genres, Freq), Freq, fill = Freq)) + 
  geom_bar(stat = "identity") + 
    geom_text(aes(label = round(Freq, dig=2)), 
            position = position_stack(vjust = 0.5), 
            color="white", size=3) + 
  coord_flip() + 
  scale_colour_brewer(palette="Set1") + 
  labs(y = 'Frequency', x = 'Genre')

# histograms of the 18 genres based on ratings
tmp = ratings %>% 
  left_join(data.frame(MovieID = movies$MovieID, genre_matrix), 
            by = "MovieID") %>%
  select(-c("UserID", "MovieID", "Rating", "Timestamp"))

data.frame(Genres = genre_list, 
                 Popularity = as.vector(colMeans(tmp))) %>% 
  ggplot(aes(reorder(Genres, Popularity), Popularity, fill = Popularity)) + 
  geom_bar(stat = "identity") + 
  geom_text(aes(label = round(Popularity, dig=3)), 
            position = position_stack(vjust = 0.5), 
            color="white", size=3) + 
  coord_flip() + 
  labs(y = 'Popularity', x = 'Genre')

tmp = rowSums(genre_matrix)
summary(tmp)

movies[which(tmp==6), ] # Movie can belong to up to six categories in the data, which is "The Transformers" (1986) that is categorized into six genres: Action, Animation, Children’s, Sci-Fi, Thriller, and War

# range(movies$Year) % 1919 to 2000
tmp = data.frame(Year = movies$Year, genre_matrix) %>%
  group_by(Year) %>%
  summarise_all(sum)
tmp[,-1] = apply(tmp[, -1], 2, cumsum)
tmp[,-1] = tmp[,-1]/sum(tmp[nrow(tmp), -1])
print(round(tmp[nrow(tmp),-1], dig=3)) # the cumulative percentages of the 18 genres users can access from 1919 to 2000, which is stored in a 81-by-19 matrix tmp

# A graphical display of the 18 genres
tmp = reshape2::melt(tmp, id.vars="Year") 
tmp %>%
  ggplot(aes(Year, value, group = variable)) +
  geom_area(aes(fill = variable)) + 
  geom_line(aes(group = variable), position = "stack")

```
#### First Scheme: Recommend Top 8 Top-Rated Movies
* Note: "top-rated" is based on users' average rating (on a 1-5 scale). This is also the scheme we apply in our recommendation system.

```{r}
# suppose we know the user's favorite genre is Crime
genre_user = "Crime"

# find out the genre 
genre_movies = genre_com[(which(genre_com[[genre_user]]== 1)),]

# reference: exploratory data analysis I
movie_rating = ratings %>% 
  group_by(MovieID) %>%
  summarize(ratings_per_movie=n(), ave_ratings = mean(Rating)) %>%
  inner_join(genre_movies, by = 'MovieID') %>%
  inner_join(movies, by = 'MovieID')

# sort by average ratings (descending order)
scheme1 = movie_rating[order(-movie_rating$ave_ratings),]

# if less than 1,000 users give ratings to the movies, then remove it from the outcome
# scheme1 = scheme1[(scheme1$ratings_per_movie >= 1000),]

# select the first 8 rows
top_8 = scheme1[1:8, ]

# only display certain columns
top_8[ , c("MovieID", "Title", "ave_ratings", "Crime")]

### The following functions also work, which are used in making the recommendation system:

# average rating for each movie in a specific genre
allmovies_genre = function(genre) {
  movieid_genre = genre_matrix[,genre] == 1
  movies_genre = movies[movieid_genre,]
  
  ratings <- ratings %>% 
    group_by(MovieID) %>% 
    mutate(avg_ratings = round(mean(Rating), dig=4)) %>%
    select(-Rating) 
  
  avg_rating<- distinct(ratings, MovieID, avg_ratings)
  
  ret = merge(avg_rating, movies_genre) %>%
    replace(is.na(.), 0) %>% 
    mutate(avg_rating = dense_rank(desc(avg_ratings))) %>% 
    arrange(desc(avg_ratings)) %>%
    distinct()
  return(ret)
}

# top N rated movies based on the average rating
topN_movies_genre = function(top_n, genre) {
  full_movies_by_genre = allmovies_genre(genre)
  ret = full_movies_by_genre %>%
    arrange(desc(avg_ratings)) %>%
    head(top_n) %>%
    select('MovieID', 'Title','avg_ratings', 'Genres')
  return(ret)
}

# a user choosing "Crime" produces the same results:
topN_movies_genre(8, "Crime")

# a user choosing "Comedy" will see the following top 8 comedies recommended:
topN_movies_genre(8, "Comedy")
```

#### second scheme: recommend top 8 most popular movies
* Note: "popularity" is measured by the number of users who give ratings to the movies. Larger number of ratings indicates more popularity.

```{r}
#### suppose we know the user's favorite genre is Crime, we can use the previous results:
# genre_user = "Crime"
# 
# # find out the genre 
# genre_movies = genre_com[(which(genre_com[[genre_user]]== 1)),]
# 
# # reference: exploratory data analysis I
# movie_rating = ratings %>% 
#   group_by(MovieID) %>%
#   summarize(ratings_per_movie=n(), ave_ratings = mean(Rating)) %>%
#   inner_join(genre_movies, by = 'MovieID') %>%
#   inner_join(movies, by = 'MovieID')

# sort by the number of ratings (descending order)
scheme2 = movie_rating[order(-movie_rating$ratings_per_movie),]

# select the first 8 rows
top_8 = scheme2[1:8, ]

# only display certain columns
top_8[ , c("MovieID", "Title", "ratings_per_movie", "Crime")]

###### Write in functions:
allmovies_pop = function(genre, top_n) {
  movieid_genre = genre_matrix[,genre] == 1
  movies_genre = movies[movieid_genre,]
  
  ratings <- ratings %>% 
    group_by(MovieID) %>% 
    summarize(ratings_per_movie=n()) 
  
  ret = merge(ratings, movies_genre) %>%
    filter(grepl(genre_user, Genres)) %>%
    replace(is.na(.), 0) %>% 
    arrange(desc(ratings_per_movie)) %>%
    select('MovieID', 'Title', 'ratings_per_movie','Genres') %>%
    head(top_n)
  
  return(ret)
}

allmovies_pop("Crime", 8) # the same results
```

## System II: Collaborative recommendation system

### Review two collaborative recommendation algorithms: UBCF and IBCF

We will be using the data set titled "ratings", which was loaded at the beginning of the file.

```{r}
library(recommenderlab)

i = paste0('u', ratings$UserID)
j = paste0('m', ratings$MovieID)
x = ratings$Rating

tmp = data.frame(i, j, x, stringsAsFactors = T)
Rmat = sparseMatrix(as.integer(tmp$i), as.integer(tmp$j), x = tmp$x)
rownames(Rmat) = levels(tmp$i)
colnames(Rmat) = levels(tmp$j)
Rmat = new('realRatingMatrix', data = Rmat)
```

### UBCF Algorithm
```{r}
set.seed(1360)
train.id = sample(nrow(ratings), floor(nrow(ratings)) * 0.8)
train = ratings[train.id, ]
test = ratings[-train.id, ]

recommender.UBCF <- Recommender(Rmat[1:500, ], method = "UBCF",
                                parameter = list(normalize = 'center', 
                                                 method = 'Cosine', 
                                                 nn = 20))

# This may take a long time
recom = predict(recommender.UBCF,
                Rmat, type = 'ratings')
rec_list = as(recom, 'list')  # each element are ratings of that user

test.pred = test
test.pred$rating = NA

# For all lines in test file, one by one
for (u in 1:nrow(test)){
    # Read userid and movieid from columns 2 and 3 of test data
    userid = as.numeric(test$UserID[u])
    movieid = as.numeric(test$MovieID[u])
    rating = rec_list[[userid]][movieid]
    # handle missing values; 2.5 might not be the ideal choice
    test.pred$rating[u] = ifelse(is.na(rating), 2.5, rating)
}
# Calculate RMSE
sqrt(mean((test$Rating - test.pred$rating)^2))
```

### IBCF Algorithm
```{r}
recommender.IBCF <- Recommender(Rmat[1:500, ], method = "IBCF",
                                parameter = list(normalize = 'center', 
                                                 method = 'Cosine', 
                                                 k = 30))

recom = predict(recommender.IBCF,
                Rmat, type = 'ratings')
rec_list = as(recom, 'list')  # each element are ratings of that user

test.pred = test
test.pred$rating = NA

# For all lines in test file, one by one
for (u in 1:nrow(test)){
    # Read userid and movieid from columns 2 and 3 of test data
    userid = as.numeric(test$UserID[u])
    movieid = as.numeric(test$MovieID[u])
    rating = rec_list[[userid]][movieid]
    # handle missing values; 2.5 might not be the ideal choice
    test.pred$rating[u] = ifelse(is.na(rating), 2.5, rating)
}
# Calculate RMSE
sqrt(mean((test$Rating - test.pred$rating)^2))
```
```




